Governing through patronage: the bargain for education in decentralized Brazil

### 2020-09-11 paper/paper.tex
second term

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations. How can these educational outcomes be improved? Previous studies have found that individual incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy experiments.\citep{finan_personnel_2015} But how do these actors behave in the institutional status quo? Who holds power over public education, and whose interests is it meant to serve?

In this paper, I build on extensive fieldwork research across Brazil to argue that the executive-legislative bargain shapes public education, leading to its capture by local political elites. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.

To analyze how politics affect staff turnover, I propose a theoretical model of legislative vote-buying, which I apply to the Brazilian context. I test the empirical implications of the model using Brazil as a case. In line with qualitative accounts and theoretical expectations, mayors who face stronger legislative opposition resort to greater patronage. Contrary to electoral accountability frameworks \citep{ferejohn_incumbent_1986,besley_political_2003}, mayors resort to patronage in order to shore up legislative support with limited consequences. I leverage data on over 2 million school teachers and school principals to estimate the effect of share of executive seats on turnover. Different model specifications confirm that turnover increases (decreases) as the share of seats held by the mayor coalition decrease (increase).

To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians confirm that turnover has a negative impact on teachers' ability to educate students. To validate these accounts I combine administrative data on education from the \emph{Prova Brasil}, a nation-wide standardized exam designed to measure student learning. I construct multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory. Using the national school census (\emph{Censo Escolar}), I construct a school-specific turnover index that tracks year-by-year volatility in teacher staff. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.

This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}. I analyze how incentives shape political decision on how to manage local bureaucracies \citep{gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders. I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services. \citep{gottlieb_countervailing_2019, ferraz_corrupting_2012}. I also bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.

@@

Positions in the educational sector are politically valuable. As a department, it comprises over 25 percent of local public sector jobs. Due to their relatively high compensation and social status compared to other positions in the local bureaucracy, educational staff positions are particularly valuable for constituencies seeking employment. Additionally, anecdotal evidence collected during fieldwork suggests that teachers play an important role in local party networks, leading electoral campaigns and brokering votes, similar to dynamics found by \citet{oliveros_making_2016} in Argentina's campaigning teachers.

@@

This study presents two sets of estimations. The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education. To do so, I leverage the test scores made available by SAEB and SPAECE to estimate the effect of turnover on student test scores. To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample. The outcome of interest is thus the average test scores for all students in the same classroom (SAEB) or same grade (SPAECE), for a given school.

>>>>

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations. How can these educational outcomes be improved? Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors behave in the institutional status quo? Who holds power over public education, and whose interests is it meant to serve?

To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. But the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In it, the exchange of public sector appointments for political favors -- a canonical example of patronage -- was designed to grant the mayor legislative support to govern according to her policy agenda.

In this paper, I argue that public education in municipalities across Brazil have been captured by political elites, with personnel appointments into public schools and management dominated by bargains between executive and legislative branches. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.

The estimation proceeds in two parts. First, I show that political alignment between mayors and city councilors has a direct ffect on patronage. Building on a canonical model, I propose a theoretical model of legislative vote-buying from which I derive my main empirical test: mayors who face less opposition in the city council engage in less patronage. To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities. In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage. This results holds across a set of specifications and different measures of turnover.

I then show that this politically induced turnover has a negative impact on student learning. To estimate the effect of patronage on quality of education, I combine qualitative and quantitative evidence. Interviews conducted with educational bureaucrats and politicians suggested firm that turnover had a negative impact on teachers' ability to educate students, as time horizons were compressed and bonds between educators and learners were broken. To validate these accounts I combine administrative data on education from two separate student learning measures: \emph{Prova Brasil} and the \emph{SPAECE}. I construct multiple datasets to test these claims: the main specification contains over 1 million classrooms spread across the national territory. A set of estimations, combining multi-level modeling and fixed effects, provide strong evidence that teacher turnover has a negative effect on student learning.

This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}. I analyze how incentives shape political decision on how to manage local bureaucracies \citep{gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders. I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services. \citep{gottlieb_countervailing_2019, ferraz_corrupting_2012}. I bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.

@@

Positions in the educational sector are politically valuable. As a department, it comprises over 25 percent of local public sector jobs. Due to their relatively high compensation and social status compared to other positions in the local bureaucracy, educational staff positions are particularly valuable for constituencies seeking employment. Additionally, qualitative evidence collected during fieldwork suggests that teachers play an important role in local party networks, leading electoral campaigns and brokering votes, similar to dynamics found by \citet{oliveros_making_2016} in Argentina's campaigning teachers.

@@

This study presents two sets of estimations. The first one identifies the effect of bureaucratic turnover on student learning, verifying the hypothesis that teacher and school principal turnover has a negative effect on the quality of municipal education. To do so, I leverage the test scores made available by \emph{SAEB} and \emph{SPAECE} to estimate the effect of turnover on student test scores. To avoid interference, individual test scores are aggregated at the classroom level, for each 5th and 9th grade of each school contained in the sample. The outcome of interest is thus the average test scores for all students in the same classroom (SAEB) or same grade (SPAECE), for a given school.
